Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-27 article.tex
f2 is not only for the tuned version

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper, at the price of parameter
tuning.

@@

To summarize, the full signal processing pipeline of our simplified (resp.
tuned) approach is defined by the composed function $w \circ h \circ g \circ
f_1$ (resp. $f_2$). When applied to the raw signal of Figure
\ref{fig:original_curve}, it outputs the signal shown in Figure
\ref{fig:weight_curve}.

@@

This appendix is structured as follows: Section \ref{app:signal} describes the differences about the signal processing methods applied on fluorescent calcium time series; Section \ref{app:averaging} then provides a detailed explanation of the averaging approach; Section \ref{app:connectome} presents a heuristic trying to retrieve directivity from fluorescent calcium time series while Section \ref{app:results} finally present results validating some choices that have been made.

@@

\subsection{Signal processing}
\ref{sapp:signal}

In Section~\ref{sec:filter}, we introduce two composed filtering functions in
order (i) to remove noise induced by light scattering artifact, (ii) to take
into account the slow decay of fluorescence signal and (iii) to reduce the
importance of global high activity in the network. However, in order to win the
challenge, we have heavily optimized several different, but yet similar,
filtering composed function. Taken alone, those composed functions in
conjunction with partial correlation might lead to worse AUROC, but improved
whenever they are averaged.

@@

\textcolor{red}{AJ: This paragraph and table should go in a separate section}
Theses filters lead, taken individually, to worse performances  than filters
$f_1$ and $f_2$ but averaging the results of all low-pass filters leads to better ROC AUC.
\begin{table}[ht]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Averaging} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 with $f_1$, $f_2$ & 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  \textbf{0.390} &  0.385 & \textbf{0.375}  \\
 with $f_1$, $f_2$, $f_3$, $f_4$ & \textbf{0.938} & \textbf{0.936} & \textbf{0.936} & \textbf{0.932} & 0.391 & 0.389 & 0.385 & 0.374\\
\hline
\end{tabular}
\caption{Performance on \textit{normal-1,2,3,4} with partial correlation with different averaging approaches.}
\end{table}

@@

\subsection{Weighted average of partial correlation statistics}

@@

\subsection{Connectome inference from pairwise interaction statistics}
\ref{sapp:connectome}

@@

At the light of the poor results of this heuristic taken on its own summarized
on Table \ref{tab:directivity} as \textit{directivity}, we use it to make
asymmetric the degree of association matrix using partial correlation
statistics. We expect that when the heuristic is highly confident (i.e.,
$z_{i,j} \gg 0 \gg z_{j,i}$) the degree of association   of edge $(i,j)$ is
increased while the degree of association corresponding to the edge $(j,i)$ is
reduced. Even if this adjustment is subtle, it might have a slight positive
effect on the ROC AUC.

@@

\subsection{Experiments}
\ref{sapp:results}

>>>>

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows:
Section~\ref{sec:filter} describes the signal processing methods applied on
fluorescent calcium time series; Section\ref{sec:inference} then presents the
proposed approach and its theoretical properties; Section~\ref{sec:results}
provides an empirical analysis and comparison with other network inference
methods, while Section~\ref{sec:conclusion} finally discusses our work and
provides further research directions. Additionally,
Appendix~\ref{app:optimized} further describes, in all details, our actual
winning method, giving slightly better results than the method presented in
the paper, at the price of parameter tuning. Appendix~\ref{app:supp} provides supplementary results on other datasets.

@@

To summarize, the full signal processing pipeline of our simplified approach is defined by the composed function $w \circ h \circ g \circ
f_1$ (resp. $f_2$). When applied to the raw signal of Figure
\ref{fig:original_curve}, it outputs the signal shown in Figure
\ref{fig:weight_curve}.

@@

This appendix is structured as follows: Section~\ref{sapp:signal} describes
the differences about the signal processing methods applied on fluorescent
calcium time series; Section~\ref{sapp:averaging} then provides a detailed
explanation of the averaging approach; Section~\ref{sapp:connectome} presents
a heuristic trying to retrieve directivity from fluorescent calcium time
series while Section~\ref{sapp:results} finally present results validating
some choices that have been made.

@@

\subsection{Signal processing}
\label{sapp:signal}

In Section~\ref{sec:filter}, we introduce two composed filtering functions in
order (i) to remove noise induced by light scattering artifact, (ii) to take
into account the slow decay of fluorescence signal and (iii) to reduce the
importance of global high activity in the network. However, in order to win the
challenge, we have heavily optimized several different, but yet similar,
filtering composed function. Taken individually, those composed functions in
conjunction with partial correlation might lead to worse AUROC, but improved
whenever they are averaged (see \ref{tab:} ).

@@

@@

\subsection{Weighted average of partial correlation statistics}
\label{sapp:averaging}

@@

\subsection{Connectome inference from pairwise interaction statistics}
\label{sapp:connectome}

@@

At the light of the poor results of this heuristic taken on its own
(on Table \ref{tab:directivity} in Section~\ref{sapp:results}) as \textit{directivity}, we use it to make
asymmetric the degree of association matrix using partial correlation
statistics. We expect that when the heuristic is highly confident (i.e.,
$z_{i,j} \gg 0 \gg z_{j,i}$) the degree of association   of edge $(i,j)$ is
increased while the degree of association corresponding to the edge $(j,i)$ is
reduced. Even if this adjustment is subtle, it might have a slight positive
effect on the ROC AUC.

\subsection{Experiments}
\label{sapp:results}

Theses filters lead, taken individually, to worse performances  than filters
$f_1$ and $f_2$ but averaging the results of all low-pass filters leads to better ROC AUC.
\begin{table}[ht]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Averaging} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 with $f_1$, $f_2$ & 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  \textbf{0.390} &  0.385 & \textbf{0.375}  \\
 with $f_1$, $f_2$, $f_3$, $f_4$ & \textbf{0.938} & \textbf{0.936} & \textbf{0.936} & \textbf{0.932} & 0.391 & 0.389 & 0.385 & 0.374\\
\hline
\end{tabular}
\caption{Performance on \textit{normal-1,2,3,4} with partial correlation with different averaging approaches.}
\end{table}

@@
